Lower Bounds on the Minimax Risk for the \\ Source Localization Problem

### 2017-01-24 paper.tex
Revise the physical sensor model section.

\usepackage[utf8]{inputenc} % allow utf-8 input
\usepackage[T1]{fontenc}    % use 8-bit T1 fonts
\usepackage{cite}
\usepackage[pdftex]{graphicx}
\usepackage{amsmath}
\usepackage{amsfonts}
\usepackage{amssymb}
\usepackage{mathtools}
\usepackage{array}
\usepackage{enumitem}
\usepackage{color}
\usepackage{url}
\usepackage[colorlinks=true,allcolors=black,urlcolor=blue]{hyperref}

@@

\author{
	\IEEEauthorblockN{
		Praveen Venkatesh\IEEEauthorrefmark{1}
		and Pulkit Grover\IEEEauthorrefmark{2}
	}
	\IEEEauthorblockA{
		Electrical \& Computer Engineering,
		and the Center for the Neural Basis of Cognition,
		Carnegie Mellon University \\
		\IEEEauthorrefmark{1}\href{mailto:vpraveen@cmu.edu}{\texttt{vpraveen@cmu.edu}}
		\IEEEauthorrefmark{2}\href{mailto:pulkit@cmu.edu}{\texttt{pulkit@cmu.edu}}
	}
}

@@

To be considered for the 2017 IEEE Jack Keil Wolf ISIT Student Paper Award.

The ``source localization'' problem is one in which we estimate the location of
a point source observed through a diffusive medium using an array of sensors.

@@

\section{Minimax lower bounds for the one-dimensional model}
\label{sec:minimax-lower-bounds}

@@

To derive the approximation, we observe that for sufficiently large numbers of
sensors, the optimal value of $\delta$ starts becoming small, as hypotheses
must chosen to be closer in order to make the error probability large. Hence,
neglecting edge effects and terms of order $\delta^3$, we can give an
approximate bound for large $m$ using a heuristic: tighten the bound by
choosing $\sqrt{\frac{2n\kappa^2\delta^2wm}{\sigma^2 S}} = \frac{1}{2}$. This
is achieved for $\delta = \sqrt{\frac{\sigma^2 S}{8nm\kappa^2 w}}$, so that for
$\Phi(\delta) = \delta^2$, equation~\eqref{eq:lb-after-delta-bound} becomes
\begin{equation}
	\mathfrak{M}_n(\mathcal{P}, \Phiorho) \geq \frac{1}{32} \frac{\sigma^2 S}{nm\kappa^2 w}.
\end{equation}
This completes the proof.
\end{IEEEproof}

@@

The bound in Theorem~\ref{thm:main-lb} may mislead one to think that the error
in source localization goes to zero, as the number of sensors goes to infinity.
This observation is qualitatively different from the information-theoretic
lower bound given in~\cite{Grover2016Fundamental}, where even with an infinite
number of sensors, the lower bound is strictly greater than zero. This
indicates that the bound in section~\ref{sec:lecam-lb} is failing to capture
some effect which leads to saturation of error, as the number of sensors
increases.

Indeed, the bound in Theorem~\ref{thm:main-lb} assumes that an arbitrarily
large number of sensors can be squeezed into a finite amount of space, and that
these sensors \emph{still} give us i.i.d.\ observations, with the same noise
variance. This is aphysical, since in reality, sensors have a certain width,
and the signal measured by them is the integral of the continuous-space signal
within that width (analogous to a sample-and-hold circuit, which integrates a
continuous-\emph{time} signal over a short period).  If the width of a sensor
is decreased, then effectively, its signal-to-noise ratio reduces. This is
intuitively evident in the sample-and-hold analogy: to compute the signal
value, the integral must be divided by the time interval over which the
integration is performed; this division amplifies noise as the time interval
gets smaller.

We performed a preliminary study to understand the effect of viewing sensors
as integrators. In the new sensor model, sensors have a finite width, which
decreases as the number of sensors increases, in order to allow their number to
grow unbounded. Concretely, the $k$-th sensor which is located at $s = kS/m$
extends from $(k{-}\frac{1}{2})S/m$ to $(k{+}\frac{1}{2})S/m$. The signal
sensed by this sensor is the $k$-th element of $\v x(\theta)$, now given by
\begin{equation}
	x_k(\theta) = \int_{(k-\frac{1}{2})S/m}^{(k+\frac{1}{2})S/m} x(s;\theta) ds
\end{equation}
The noise vector $\v \epsilon$ is also generated similarly, by integrating over
small sections of a wide-sense-stationary white noise process, which is assumed
to additively corrupt the signal. This leads to i.i.d.\ noise values at sensors
(since they do not overlap), with variance inversely proportional to $m$, i.e.,
$\sigma(m) \propto 1/\sqrt{m}$.

The bounding technique used in section~\ref{sec:lecam-lb} no longer easily
admits a solution which gives a bound for all values of $m$. We can compute the
bound numerically, however, using exact expressions for error probability in a
binary hypothesis test. For gaussian distributions of equal variance and a
uniform prior on the hypotheses, the best possible error probability (over all
tests) is a Q-function~\cite{Qfunction} in the distance between their
means~\cite{NPLemmaOrSimilar},
\begin{equation}
	P_e = \inf_\psi \mathbb P(\psi(Y^n) \neq V) = Q\bigg(\frac{\norm{\v\mu_0 - \v\mu_1}}{2\sigma(m)}\bigg).
\end{equation}
The minimax estimation risk is lower bounded by the risk in hypothesis testing,
hence
\begin{equation}
	\mathfrak{M}_n(\mathcal{P}, \Phiorho) \geq \Phi(\delta) P_e = \delta^2 Q\bigg(\frac{\norm{\v x(\theta_0) - \v x(\theta_1)}}{2\sigma(m)}\bigg).
\end{equation}
This bound is valid for any pair of hypotheses $\theta_0$ and $\theta_1$,
separated by a distance of $2\delta$. As the $\delta$ increases, it becomes
easier to distinguish between the two hypotheses, so $P_e$ falls. When
numerically evaluating the bound, therefore, we have to \emph{manually}
trade-off between $\delta$ and $P_e$ to find the tightest possible bound. This
comes down to choosing $\theta_0$ and $\theta_1$ so that they are the hardest
hypotheses to distinguish between. To make the simulation easier, we can relax
the process of finding the least discernable hypothesis pair by fixing
$\theta_0$. Then, we need only optimize over $\delta$, making the process more
tractable. This comes at the cost of loosening the lower bound.

@@

Plotting this numerically computed bound for different numbers of trials (see
Fig.~\ref{fig:numerical}) reveals some interesting trends. First, the bound
saturates as $m$ grows large, matching the behaviour presented
in~\cite{Grover2016Fundamental}. Second, the lower bound on error is larger for
smaller numbers of sensors, as we expect. With fewer sensors, it is harder to
estimate the location of the point source. Third and most importantly, the
point at which the bound begins to saturate increases when we use
more trials. Recall that the number of ``trials'' is the number of i.i.d.\
sensor observations we receive. Intuitively, estimation is eased with more
trials, as we could perform more averaging, and increase SNR. \emph{With higher
SNR, there is greater benefit in using a larger number of sensors.} This
matches the intuition given in~\cite{Grover2016Information}, where it is
propounded that using more sensors allows us to sample a greater number of
high-frequency components of the source signal, and hence get a better estimate
of its location.

@@

While the bounds in section~\ref{sec:lecam-lb} were derived for a
one-dimensional circular domain, extensions to a one-dimensional linear domain
(of fixed length) and to a $d$-dimensional domain (circular or linear) are
fairly straightforward, under certain conditions. For a 1D linear domain, we
require that sensors capture the entirety of the impulse response even at the
edges of the domain. Hence, sensors must be placed uniformly in $[-w/2,
S{+}w/2]$. For extending to $d$-dimensions, it suffices to apply Assouad's
Lemma~\cite{Tsybakov2009Introduction} to separate the problem into $d$
independent 1D problems.

@@

Another point on sensor placement bears repeating: the bounds derived in this
paper apply to only a specific sensor configuration, i.e., the uniform one.
This would intuitively seem to be an appropriate choice in the minimax setting,
however, since any non-uniform placement will have one or more points which is
far from all sensors. The ``maximization'' over $\theta \in \Theta$ would then
select a $\theta$ for which the value of the impulse response at the sensors is
smallest, so that SNR is minimized.

>>>>

\usepackage[utf8]{inputenc} % allow utf-8 input
\usepackage[T1]{fontenc}    % use 8-bit T1 fonts
\usepackage{cite}
\usepackage[pdftex]{graphicx}
\usepackage{amsmath}
\usepackage{amsfonts}
\usepackage{amssymb}
\usepackage{mathtools}
\usepackage{nicefrac}
\usepackage{array}
\usepackage{enumitem}
\usepackage{color}
\usepackage{url}
\usepackage[colorlinks=true,allcolors=black,urlcolor=blue]{hyperref}

@@

\author{
	\IEEEauthorblockN{
		\textcolor{blue}{Praveen Venkatesh %\IEEEauthorrefmark{1}
		and Pulkit Grover} %\IEEEauthorrefmark{2}
	}
	%\IEEEauthorblockA{
	%    Electrical \& Computer Engineering,
	%    and the Center for the Neural Basis of Cognition,
	%    Carnegie Mellon University
	%    \\
	%    \IEEEauthorrefmark{1}\href{mailto:vpraveen@cmu.edu}{\texttt{vpraveen@cmu.edu}}
	%    \IEEEauthorrefmark{2}\href{mailto:pulkit@cmu.edu}{\texttt{pulkit@cmu.edu}}
	%}
}

@@

To be considered for the 2017 IEEE Jack Keil Wolf ISIT Student Paper Award.
The ``source localization'' problem is one in which we estimate the location of
a point source observed through a diffusive medium using an array of sensors.

@@

\section{Lower bounds on the minimax risk for the one-dimensional model}
\label{sec:minimax-lower-bounds}

@@

To derive the approximation, we observe that for sufficiently large numbers of
sensors, the optimal value of $\delta$ starts becoming small, as hypotheses
must chosen to be closer in order to make the error probability large. Hence,
neglecting edge effects and terms of order $\delta^3$, we can give an
approximate bound for large $m$ using a heuristic: tighten the bound by
choosing $\sqrt{\frac{2n\kappa^2\delta^2wm}{\sigma^2 S}} = \frac{1}{2}$. This
is achieved for $\delta = \sqrt{\frac{\sigma^2 S}{8nm\kappa^2 w}}$, so that for
$\Phi(\delta) = \delta^2$, equation~\eqref{eq:lb-after-delta-bound} becomes
\begin{equation}
	\mathfrak{M}_n(\mathcal{P}, \Phiorho) \gtrsim \frac{1}{32} \frac{\sigma^2 S}{nm\kappa^2 w}.
\end{equation}
This completes the proof.
\end{IEEEproof}

@@

The bound in Theorem~\ref{thm:main-lb} may mislead one to think that the error
in source localization goes to zero, as the number of sensors goes to infinity.
\textcolor{blue}{
This observation is qualitatively different from the information-theoretic
lower bound given in~\cite{Grover2016Fundamental}, where even with an infinite
number of sensors, the lower bound is strictly greater than zero. This
indicates that the bound in section~\ref{sec:lecam-lb} is failing to capture
some effect which leads to saturation of error, as the number of sensors
increases.
}

Indeed, the bound in Theorem~\ref{thm:main-lb} assumes that an arbitrarily
large number of sensors can be squeezed into a finite amount of space, and that
these sensors \emph{still} give us i.i.d.\ observations, with the same noise
variance. This is aphysical, since in reality, each sensor has a certain width,
and the value measured by it is the integral of the continuous-space signal
within that width (analogous to a sample-and-hold circuit, which integrates a
continuous-\emph{time} signal over a short period).  If the width of a sensor
is decreased, then effectively, its signal-to-noise ratio reduces.  This is
intuitively evident in the sample-and-hold analogy: to compute the signal
value, the integral must be divided by the time interval over which the
integration is performed; this division amplifies noise as the time interval
gets smaller.

We performed a preliminary study to understand the effect of viewing sensors as
integrators. In the new sensor model, sensors have a width which decreases as
their number grows to infinity. Concretely, the $k$-th sensor which is located
at $s = kS/m$ extends from $(k{-}\frac{1}{2})S/m$ to $(k{+}\frac{1}{2})S/m$.
The signal sensed by this sensor is the $k$-th element of $\v x(\theta)$, now
given by
\begin{equation}
	x_k(\theta) = \int_{(k-\frac{1}{2})S/m}^{(k+\frac{1}{2})S/m} x(s;\theta) ds
\end{equation}
\textcolor{blue}{
The noise vector $\v \epsilon$ is also generated similarly, by integrating over
small sections of a wide-sense-stationary white noise process, which is assumed
to additively corrupt the continuous-space signal $x(s;\theta)$ prior to
sampling. This leads to i.i.d.\ noise values at sensors (since they do not
overlap), with variance inversely proportional to $m$, i.e., $\sigma(m) \propto
1/\sqrt{m}$.
}

The bounding technique used in section~\ref{sec:lecam-lb} no longer easily
admits a solution which gives a bound for all values of $m$. We instead compute
the bound numerically, using exact expressions for error probability in a
binary hypothesis test. For gaussian distributions of equal variance and a
uniform prior on the hypotheses, the minimum possible error probability (over
all tests) is a Q-function in the distance between their
means~\cite{NPLemmaOrSimilar},
\begin{equation}
	P_e = \inf_\psi \mathbb P(\psi(Y^n) \neq V) = Q\bigg(\frac{\norm{\v\mu_0 - \v\mu_1}}{2\sigma(m)}\bigg).
\end{equation}
where $Q(x) = \int_x^\infty \frac{1}{\sqrt{2\pi u}} e^{-u^2/2} du$. Hence, from
Theorem~\ref{thm:est-to-testing},
\begin{equation}
	\mathfrak{M}_n(\mathcal{P}, \Phiorho) \geq \Phi(\delta) P_e = \delta^2 Q\bigg(\frac{\norm{\v x(\theta_0) - \v x(\theta_1)}}{2\sigma(m)}\bigg).
\end{equation}
This bound is valid for any $\{\theta_0, \theta_1\}$ constituting a
$2\delta$-packing of $\Theta$. To numerically optimize over all possible
packings, we relax the process by fixing $\theta_0 = 0$ and optimizing over
$\theta_1 = 2\delta$.  We compute sensor values by performing numerical
integration, and find the best possible $\delta$ by grid search.

%This bound is valid for any pair of hypotheses $\theta_0$ and $\theta_1$,
%separated by a distance of $2\delta$.
%As $\delta$ increases, it becomes
%easier to distinguish between the two hypotheses, so $P_e$ falls. When
%numerically evaluating the bound, therefore, we have to \emph{manually}
%trade-off between $\delta$ and $P_e$ to find the tightest possible bound. This
%comes down to choosing $\theta_0$ and $\theta_1$ so that they are the hardest
%hypotheses to distinguish between. To make the simulation easier, we can relax
%the process of finding the least discernable hypothesis pair by fixing
%$\theta_0$. Then, we need only optimize over $\delta$, making the process more
%tractable. This comes at the cost of loosening the lower bound.

@@

Plotting this numerically computed bound for different numbers of trials (see
Fig.~\ref{fig:numerical}) reveals some interesting trends. First, the bound
saturates as $m$ grows large, matching the behaviour presented
in~\cite{Grover2016Fundamental}. Second, the lower bound on error is larger for
smaller numbers of sensors, as we expect. Third and most importantly, as the
number of trials increases, bound begins to saturate only for larger numbers of
sensors. In other words, there is increased benefit in using larger numbers of
sensors, when we also have more trials\footnote{Recall that the number of
``trials'' is the number of i.i.d.\ sensor observations we receive, per
equation~\ref{eq:sensor-obs}.}. This substantiates the intuition given
in~\cite{Grover2016Information}, which addresses this issue using a
Nyquist-rate perspective. A larger number of sensors will sample a larger
number of high-frequency components of the source signal (and hence allow us to
locate it better), but only if the noise floor is sufficiently low. Noise
variance, in turn, can be reduced by averaging over multiple trials.

@@

While the bounds in section~\ref{sec:lecam-lb} were derived for a
one-dimensional circular domain, extensions to a one-dimensional linear domain
(of fixed length) and to a $d$-dimensional domain (circular or linear) are
fairly straightforward, under certain conditions. For a 1D linear domain, we
require that sensors capture the entirety of the impulse response even at the
edges of the domain. Hence, sensors must be placed uniformly in $[-w/2,
S{+}w/2]$. For extending to $d$ dimensions, it suffices to apply Assouad's
Lemma~\cite{Tsybakov2009Introduction} to separate the problem into $d$
independent one-dimensional problems.

Another point on sensor placement bears repeating: the bounds derived in this
paper apply to only a specific sensor configuration, i.e., the uniform one.
Intuitively, this is the correct choice in the minimax setting, since any
non-uniform placement will have one or more points which is far from all
sensors. The ``maximization'' over $\theta \in \Theta$ would then select a
$\theta$ for which the value of the impulse response at the sensors is
smallest, so that the ratio of mean-sensor-value to noise variance (a form of
``SNR'') is minimized.

@@
